---
title: "Analysis of honey bee wing shape in Algeria"
subtitle: ""
author: 
  - name: "Yamina Haider"
    affiliation: University of Boumerdes, Algeria
    orcid: 0000-0001-6085-3513
  - name: "Adam Tofilski"
    affiliation: University of Agriculture in Krakow, Poland
    orcid: 0000-0002-3898-7029
date: today
date-format: iso
format: 
  html:
    self-contained: true
    toc: true
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  tidy = TRUE, tidy.opts = list(arrow=TRUE, indent=2)
)
options(digits=3) 

setwd("C:/my_programs/GitHub/Apis-mellifera-DZ")
```

## Libraries

```{r libraries}
# calculations
library(geomorph) # GPA 
library(Morpho) # CVA
library(IdentiFlyR)  # classification
library(phangorn) #UPGMA

# plotting and visualization
library(ggplot2) # plots
ggplot2::theme_set(theme_light())
library(rnaturalearth) # maps
library(terra) # raster maps
library(ggpubr) # ggarrange

```

## Variable names

```{r var-names}

p = 19 # number of landmarks
k = 2  # number of dimensions, in this case 2 for coordinates (x, y)

# create coordinates names used by IdentiFly
xyNames = c("x1", "y1")
for (i in 2:p) {
  xyNames = c(xyNames, paste0("x", i))
  xyNames = c(xyNames, paste0("y", i))
}
xyNames

# The number of principal components used is 2*p-4 = 34, which is equal to the degrees of freedom
# create principal components names used by prcomp
pcNames <- paste0("PC", 1:(2*p-4))
pcNames

# define which landmarks are connected by lines in wireframe graph
link.x <- c(1, 1, 2, 2, 3,  3, 4,  4,  5, 6, 7,  7,  7, 8,  9,  9, 10, 11, 11, 12, 13, 14, 15, 16, 17)
link.y <- c(2, 3, 4, 5, 6, 19, 6, 10, 12, 8, 8, 14, 19, 9, 10, 15, 11, 12, 16, 13, 18, 15, 16, 17, 18)
links.apis <- cbind(link.x, link.y)

GP1 <- gridPar(pt.bg = "blue", link.col = "blue", pt.size = 1, tar.pt.bg = "red",
  tar.link.col = "red")

pointSize = 1

```

## Read data
We used following datasets available at Zenodo: <a href="#Haider & Tofilski, 2024">Haider & Tofilski, 2024</a>, <a href="#Haider & Tofilski, 2026">Haider & Tofilski, 2026</a>, <a href="#Tofilski, 2026">Tofilski, 2026</a>
As references samples of *A. m. intermissa* and *A. m. sahariensis* we used supplementary data provided by <a href="#Bakhchou et al., 2025">Bakhchou et al., 2025</a>

```{r read-coordinates}

wings = read.csv("https://zenodo.org/records/18360081/files/DZ-2025-raw-coordinates.csv")
wings <- rbind(wings, read.csv("https://zenodo.org/records/18359322/files/DZ-2018-raw-coordinates.csv"))
wings <- rbind(wings, read.csv("https://zenodo.org/records/13939415/files/DZ-2024-raw-coordinates.csv"))

geoData = read.csv("https://zenodo.org/records/18360081/files/DZ-2025-data.csv")
geoData <- rbind(geoData, read.csv("https://zenodo.org/records/18359322/files/DZ-2018-data.csv"))
geoData2024 = read.csv("https://zenodo.org/records/13939415/files/DZ-2024-data.csv")
geoData2024$notes = "Mediterranean"
geoData <- rbind(geoData, geoData2024)

names(geoData)[names(geoData) == "notes"] <- "climate"
geoData$climate <- factor(geoData$climate, levels = c("desert", "semiarid", "Mediterranean"))  #specify order

# read reference of A. m. intermissa and A. m. sahariensis
dataMA = read.csv("MA.csv", sep = ";")
# use only A. m. intermissa and A. m. sahariensis
dataMA = dataMA[dataMA$notes %in% c("A. m. intermissa", "A. m. sahariensis"),]

wingsMA = dataMA[,c("file", xyNames)]
wings = rbind(wings, wingsMA)

geoDataMA = dataMA[,c("file", "sample", "latitude", "longitude", "date")]

geoDataMA = data.frame(file = dataMA$file,
                       sample = dataMA$sample,
                       latitude = dataMA$latitude,
                       longitude = dataMA$longitude,
                       date = dataMA$date,
                       resolution = NA,
                       climate = dataMA$notes)
geoDataDZ = geoData
geoData = rbind(geoData, geoDataMA)

```

## GPA

```{r GPA}

# Convert from 2D array to 3D array
wings3D <- arrayspecs(wings[xyNames], p, k)
dimnames(wings3D)[[3]] <- wings$file

# Align the coordinates using Generalized Procrustes Analysis 
GPA <- gpagen(wings3D, print.progress = FALSE)

# plot landmarks after alignment
plotAllSpecimens(GPA$coords,links=links.apis,label=TRUE,
                 plot_param = list(pt.bg = "black", pt.cex = 0.5,
                                   mean.bg = "red", mean.cex=1, link.col="red",
                                   txt.pos=3, txt.cex=1))

# Convert from 3D array to 2D array
wingsAligned <- two.d.array(GPA$coords) 
colnames(wingsAligned) = xyNames
wingsAligned = as.data.frame(wingsAligned)

samples <- aggregate(wingsAligned, by = list(geoData$sample), FUN = mean)
samples <- data.frame(samples, row.names = 1) # move column 1 to row names

geoDataSample <- aggregate(geoData[c("latitude", "longitude", "date")], by = list(geoData$sample), FUN = mean)
gr = aggregate(climate ~ sample, data = geoData, FUN = unique)
geoDataSample$climate = gr$climate

geoDataSample <- data.frame(geoDataSample, row.names = 1) # move column 1 to row names

samples = cbind(samples, geoDataSample)
samples$climate <- factor(samples$climate, levels = c("desert", "semiarid", "Mediterranean"))  #specify order
samplesAll = samples
samples = samplesAll[samplesAll$climate %in% c("desert", "semiarid", "Mediterranean"),]

```

## Sample size

```{r n}

# number of wings
nrow(geoDataDZ)
# number of wings per species
table(geoDataDZ$climate)

# number of samples
nrow(samples)
# number of samples per species
table(samples$climate)

```

## Maps

```{r map}

# Read elevation data
elev <- rast("X:/WorldClim/wc2.1_2.5m_elev.tif")

colorElev <- colorRampPalette(c("#f7f7f7", "#f0f0f0", "#d9d9d9", "#bdbdbd", "#636363"), bias=3)

xMin = -10
xMax = 15
yMin = 19
yMax = 40
e <- ext(xMin - 5, xMax + 5, yMin - 5, yMax + 5) 
elev = crop(elev, e)
elevDf <- as.data.frame(elev, xy = TRUE, na.rm = TRUE)
colnames(elevDf) <- c("longitude", "latitude", "altitude")

world <- ne_countries(scale = "medium", returnclass = "sf")

ggplot(data = world) +
  geom_raster(data = elevDf, aes(longitude, latitude, fill = altitude)) +
  scale_fill_gradientn(
    colours = colorElev(100),
    guide = guide_colorbar(order = 3)
  ) +
  geom_sf(fill = NA) +
  coord_sf(xlim = c(xMin, xMax), ylim = c(yMin, yMax)) +
  geom_jitter(data = samples, aes(x = longitude, y = latitude, color = climate), size = pointSize, width = 0.1, height = 0.1) +
  scale_color_manual(
    values = c("red", "yellow", "green"),
    guide = guide_legend(order = 1)
  ) +
  theme(legend.position="right") 

```

## Classification as lineages
Comparison with lineages dataset <a href="#Nawrocka et al., 2018a">Nawrocka et al., (2018a)</a>, <a href="#Nawrocka et al., 2018b">Nawrocka et al., (2018b)</a>

```{r class-lin}
idData = xml2gmLdaData("https://zenodo.org/records/14054009/files/apis-mellifera-lineage.dw.xml")

id = gmLdaData2id(idData,
                  samples[, xyNames],
                  average = FALSE)

id$plot
table(id$id$group)
pTab = as.data.frame(id$P)

# colony with minimum probability of classification as lineage A
pTab[which.min(pTab$A), ]
# colony with maximum probability of classification as lineage C
pTab[which.max(pTab$C), ]
# colony with maximum probability of classification as lineage M
pTab[which.max(pTab$M), ]
# colony with maximum probability of classification as lineage O
pTab[which.max(pTab$O), ]

```

## PCA

```{r PCA}

PCA <- prcomp(samples[ , xyNames])
PCscores <- as.data.frame(PCA$x)
PCscores <- cbind(PCscores, samples[, c("latitude", "longitude", "climate")])

# create plot labels
variance.tab <- summary(PCA)$importance
variance <- variance.tab["Proportion of Variance", "PC1"]
variance <- round(100 * variance, 1)
label.x <- paste0("PC1 (", variance, "%)")
variance <- variance.tab["Proportion of Variance", "PC2"]
variance <- round(100 * variance, 1)
label.y <- paste0("PC2 (", variance, "%)")

grN = 3
figPca = ggplot(PCscores, aes(x = PC1, y = PC2, color = climate)) +
  geom_point() +
  scale_color_manual(name = "", values = rainbow(grN)) +
  stat_ellipse() + 
  xlab(label.x) + ylab(label.y)
figPca

```

## Differences between climates

```{r manova}

samples3D <- arrayspecs(samples[, xyNames], p, k)
gdf <- geomorph.data.frame(coords = samples3D, 
                           climate = samples$climate)
gdf$climate <- factor(gdf$climate)


fit <- procD.lm(
  coords ~ climate,
  data = gdf
)
summary(fit)

PW <- pairwise(fit, groups = gdf$climate)
# distances among means
summary(PW, confidence = 0.95, test.type = "dist") 
# correlation between mean vectors (angles in degrees)
summary(PW, confidence = 0.95, test.type = "VC", 
   angle.type = "deg")

# desert - Mediteranean
DZde_mean <- mshape(samples3D[,, which(samples$climate == "desert")])
DZme_mean <- mshape(samples3D[,, which(samples$climate == "Mediterranean")])
plotRefToTarget(DZde_mean, DZme_mean,
                mag=4, links=links.apis, gridPars = GP1, method = "points")
legend(
  "bottomleft",
  legend = c("desert", "Mediteranean"),
  pch = 21,
  pt.bg = c("blue", "red"),
  bty = "n"
)

# desert-semiarid
DZse_mean <- mshape(samples3D[,, which(samples$climate == "semiarid")])
plotRefToTarget(DZde_mean, DZse_mean,
                mag=4, links=links.apis, gridPars = GP1, method = "points")
legend(
  "bottomleft",
  legend = c("desert", "semiarid"),
  pch = 21,
  pt.bg = c("blue", "red"),
  bty = "n"
)

# semiarid-Mediteranean
plotRefToTarget(DZse_mean, DZme_mean,
                mag=4, links=links.apis, gridPars = GP1, method = "points")
legend(
  "bottomleft",
  legend = c("semiarid", "Mediteranean"),
  pch = 21,
  pt.bg = c("blue", "red"),
  bty = "n"
)

```


## LDA climate

```{r LDA}
# use equal prior probability for all groups
grN <- length(unique(PCscores$climate)) # number of groups
sample.cva <- CVA(PCscores[pcNames], PCscores$climate,
                  rounds = 10000, cv = TRUE,
                  prior = rep(1/grN, grN))
sample.cva.scores <- as.data.frame(sample.cva$CVscores)
# rename variable names from CV to LD otherwise use `CV 1`
colnames(sample.cva.scores) <- gsub("CV ", "LD", colnames(sample.cva.scores)) 
# sample.cva.scores <- cbind(sample.geoData, sample.cva.scores)
sample.cva.scores$climate = PCscores$climate

figLda = ggplot(sample.cva.scores, aes(x = LD1, y = LD2, color = climate)) +
  geom_point() +
  scale_color_manual(name = "", values = rainbow(grN)) +
  stat_ellipse()
figLda

ggarrange(figPca,
          figLda,
          labels = c("a", "b"),
          font.label = list(size = 12, face = "bold"),
          ncol = 2, nrow = 1, common.legend = TRUE, legend="bottom")

# Confusion matrix
# Classification of the samples to groups
LDA.class <- typprobClass(sample.cva$CVscores, groups = as.factor(sample.cva.scores$climate), outlier = 0)
print(LDA.class)

# Mahalanobis distances between groups
knitr::kable(as.data.frame(as.matrix(sample.cva$Dist$GroupdistMaha)), digits = 3)
# pairwise probabilities of differences between groups
knitr::kable(as.data.frame(as.matrix(sample.cva$Dist$probsMaha)), digits = 5)

MD = as.matrix(sample.cva$Dist$GroupdistMaha)
tree = upgma(MD)
plot(tree)

```

## LDA subspecies

```{r LDA-sub}
PCA <- prcomp(samplesAll[ , xyNames])
PCscores <- as.data.frame(PCA$x)
PCscores$gr = geoDataSample$climate

# use equal prior probability for all groups
grN <- length(unique(PCscores$gr)) # number of groups
sample.cva <- CVA(PCscores[pcNames], PCscores$gr,
                  rounds = 10000, cv = TRUE,
                  prior = rep(1/grN, grN))
sample.cva.scores <- as.data.frame(sample.cva$CVscores)
# rename variable names from CV to LD otherwise use `CV 1`
colnames(sample.cva.scores) <- gsub("CV ", "LD", colnames(sample.cva.scores)) 
# sample.cva.scores <- cbind(sample.geoData, sample.cva.scores)
sample.cva.scores$gr = PCscores$gr

ggplot(sample.cva.scores, aes(x = LD1, y = LD2, color = gr)) +
  geom_point() +
  scale_color_manual(name = "", values = c(rainbow(3), "gray", "black")) +
  stat_ellipse()

# Confusion matrix
# Classification of the samples to groups
LDA.class <- typprobClass(sample.cva$CVscores, groups = as.factor(sample.cva.scores$gr), outlier = 0)
print(LDA.class)

# Mahalanobis distances between groups
knitr::kable(as.data.frame(as.matrix(sample.cva$Dist$GroupdistMaha)), digits = 3)
# pairwise probabilities of differences between groups
knitr::kable(as.data.frame(as.matrix(sample.cva$Dist$probsMaha)), digits = 5)

```

## References

<p id="Bakhchou et al., 2025">
Bakhchou, S., Aglagane, A., Tofilski, A., Mokrini, F., Er-Rguibi, O., El Mouden, E. H., Machlowska, J., Fellahi, S., & Mohssine, E. H. (2025). Morphological Diversity of Moroccan Honey Bees (*Apis mellifera* L. 1758): Insights from a Geometric Morphometric Study of Wing Venation in Honey Bees from Different Climatic Regions. Diversity, 17(8), 527. <https://doi.org/10.3390/d17080527>
</p>
<p id="Haider & Tofilski, 2024">
Haider, Y., & Tofilski, A. (2024). Fore wing images of honey bees (*Apis mellifera*) from Algeria 2024 [Dataset]. Zenodo. <https://doi.org/10.5281/zenodo.13939415>
</p>
<p id="Haider & Tofilski, 2026">
Haider, Y., & Tofilski, A. (2026). Fore wing images of honey bees (*Apis mellifera*) from Algeria 2025 [Dataset]. Zenodo. <https://doi.org/10.5281/zenodo.18360081>
</p>
<p id="Nawrocka et al., 2018a">
Nawrocka, A., Kandemir, İ., Fuchs, S., & Tofilski, A. (2018). Computer software for identification of honey bee subspecies and evolutionary lineages. Apidologie, 49(2), 172-184. <https://doi.org/10.1007/s13592-017-0538-y>
</p>
<p id="Nawrocka et al., 2018b">
Nawrocka, A., Kandemir, İ., Fuchs, S., & Tofiilski, A. (2018). Dataset: Computer software for identification of honey bee subspecies and evolutionary lineages. Apidologie, 49, 172–184. <https://doi.org/10.5281/zenodo.7567336>
</p>
<p id="Tofilski, 2026">
Tofilski, A. (2026). Fore wing images of honey bees (*Apis mellifera*) from Algeria 2018 [Dataset]. Zenodo. <https://doi.org/10.5281/zenodo.18359322>
</p>

## Session information

```{r info}

sessionInfo()

```
